{\huge \textbf{Unshuffling Permutations}}

### 2016-04-09 slides/slides.tex
Ameliorations slides

%\usepackage{euler,palatino}
\usepackage{lmodern}
\usepackage{amsmath,amssymb}
\usepackage[utf8]{inputenc}
\usepackage{xxcolor}
\usepackage{graphicx}
\usepackage[ruled]{algorithm2e}
\usepackage[english]{babel}
\usepackage{rotate}
\usepackage{shuffle}
%\usepackage{minted}
\usepackage{url}
\usepackage{etoolbox}
\usepackage{tikz}
\usepackage{rotating}
\usetikzlibrary{snakes}
\usetikzlibrary{trees}
\usetikzlibrary{arrows}
\usetikzlibrary{matrix}
\usetikzlibrary{calc}

@@

\title[Unshuffling Permutations]{%
  \huge \textbf{Unshuffling Permutations}
  \vspace{1.5em}
}

\subtitle{
    \textbf{\underline{Samuele Giraudo}}\qquad\textbf{St\'ephane~Vialette}\\[1em]
    {\footnotesize Laboratoire d'Informatique Gaspard-Monge\\
    Université Paris-Est Marne-la-Vallée \\
    UMR CNRS 8049 \\[2em]}
    \includegraphics[scale=.16]{logos/CNRSfr}
    \qquad
    \includegraphics[scale=.10]{logos/ecole_ponts_RVB300.jpg}
    \qquad
    \includegraphics[scale=.26]{logos/LIGM}
    \qquad
    \raisebox{.25cm}{\includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/LogoEsieeParis}}
    \qquad
    \raisebox{.25cm}{\includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/UPEM_LOGO_EDITION}}
}

\date[LATIN 16]{
    {
    \scriptsize Latin American Theoretical Informatics Symposium \\
    April $11$-$15$ $2016$
    }
}% end date

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Shuffling words}
\begin{block}{Definition}
    The \alert{shuffle product} $\shuffle$ on words of $A^*$ is defined
    recursively by \vspace{-.5em}
    \begin{align*}
        u \shuffle \epsilon & := \{u\} =: \epsilon \shuffle u, \\
        (ua \shuffle vb) & := (ua \shuffle v)b \cup (u \shuffle vb)a.
    \end{align*}
\end{block}
\medskip

\uncover<2->{%
To take into account multiplicities, we consider $\shuffle$ as a linear
product
\begin{equation*}
    \shuffle : \QQ[A^*] \otimes \QQ[A^*] \to \QQ[A^*]
\end{equation*}
on $\QQ[A^*]$, the $\QQ$-linear span of words defined by
\begin{align*}
    u \shuffle \epsilon & := u =: \epsilon \shuffle u, \\
    (ua \shuffle vb) & := (ua \shuffle v)b + (u \shuffle vb)a.
\end{align*}}
%
\uncover<3->{%
\begin{block}{Example}
    \vspace{-1em}
    \begin{equation*}\begin{split}
        \textcolor{RoyalBlue}{\La\Lb} \shuffle \textcolor{BrickRed}{\Lb\La}
        & =
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\Lb\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb}%
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb\La}%
        \textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La\Lb}%
        \textcolor{BrickRed}{\La}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La}%
        \textcolor{BrickRed}{\La}\textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb\La}\textcolor{RoyalBlue}{\La\Lb} \\
        &
        \uncover<4->{=
        2 \La\Lb\Lb\La + \La\Lb\La\Lb + \Lb\La\Lb\La + 2 \Lb\La\La\Lb}
    \end{split}\end{equation*}
    \vspace{-1.5em}
\end{block}}
\end{frame}

@@

Given two words $v_1$ and $v_2$, the set $v_1 \shuffle v_2$
can be computed in
\begin{equation*}
    O\left((|v_1| + |v_2|) \; \binom{|v_1| + |v_2|}{|v_1|}\right)
\end{equation*}
time \BIB{Spehner}{1986}.
\bigskip

\uncover<2->{%
Given three words $u$, $v_1$, and $v_2$, deciding if $u$ is in
$v_1 \shuffle v_2$ can be done in
\begin{equation*}
O\left(\frac{|u|^2}{\log(|u|)}\right)
\end{equation*}
time \BIB{van Leeuwen, Nivat}{1982}.
\bigskip}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Square words}
\begin{block}{Definition}
    A word $u$ is a \alert{square} (w.r.t. the shuffle product
    $\shuffle$) if there is a word $v$ such that $u$ appears
    in~$v \shuffle v$.
\end{block}
\medskip

\uncover<2->{%
\begin{block}{Example}
    The word
    \begin{math}
        u := \textcolor{RoyalBlue}{\Lc}\textcolor{BrickRed}{\Lc}
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\La}
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\Lb\Lb}
    \end{math}
    is a square since $u$ can be obtained by shuffling $\Lc\La\Lb\Lb$
    with itself.
\end{block}
\medskip}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Recognizing square words}
\begin{block}{Definition}
    A \alert{perfect matching} of a word $u$ is a graph $(V, E)$ such
    that
    \begin{itemize}
        \item $V := \{(u_i, i) : i \in \{1, \dots, |u|\}$;
        \item every vertex of $V$ belongs to exactly one edge of $E$;
        \item $(u_i, i) \-- (u_j, j) \in E$ implies $u_i = u_j$.
    \end{itemize}
\end{block}

\uncover<2->{%
\begin{block}{Example}
    The word $\Lb\La\Lb\La\Lb\Lb\Lc\Lc$ admits the perfect matching \vspace{-1em}
    \begin{equation*}
    \begin{split}
    \begin{tikzpicture}[xscale=.5,yscale=.3,inner sep=1pt,node distance=1.5cm]
        \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
        \draw node[Vertex,fill=RoyalBlue!30](U01){$(\Lb, 1)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(\La, 2)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(\Lb, 3)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(\La, 4)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U04](U05) {$(\Lb, 5)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U05](U06){$(\Lb, 6)$};
        \draw node [Vertex,fill=RoyalBlue!30,right of=U06](U07){$(\Lc, 7)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U07](U08){$(\Lc, 8)$};
        \draw [thick,>=latex']
        (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
            and ($ (U06.north) + (0,2.5) $) .. (U06.north);
        \draw [thick,>=latex']
        (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
            and ($ (U04.north) + (0,1.5) $) .. (U04.north);
        \draw [thick,>=latex']
        (U03.south) .. controls ($ (U03.south) + (0,-1.5) $)
            and ($ (U05.south) + (0,-1.5) $) .. (U05.south);
        \draw [thick,>=latex']
        (U07.north) .. controls ($ (U07.north) + (0,1.5) $)
            and ($ (U08.north) + (0,1.5) $) .. (U08.north);
    \end{tikzpicture}
    \end{split}\,.
    \end{equation*}
    \vspace{-1.5em}
\end{block}}

\uncover<3->{%
\begin{block}{Definition}
    A perfect matching $(V, E)$ of a word $u$ is \alert{inclusion-free}
    if there are no edges $(u_i, i) \-- (u_j, j)$ and
    $(u_k, k) \-- (u_\ell, \ell)$ of $E$ such that $i < k < \ell < j$.
\end{block}}
\end{frame}

@@

\begin{block}{Proposition \BIB{Rizzi, Vialette}{2013}}
    A word $u$ is a square iff $u$ admits an inclusion-free perfect
    matching.
\end{block}

\uncover<2->{%
\begin{block}{Example}
    The word
    \begin{math}
        \textcolor{RoyalBlue}{\Lc}\textcolor{BrickRed}{\Lc}
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\La}
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\Lb\Lb}
    \end{math}
    admits the perfect matching \vspace{-1.25em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.16,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(\Lc, 1)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U01](U02) {$(\Lc, 2)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(\La, 3)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U03](U04) {$(\Lb, 4)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U04](U05) {$(\La, 5)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U05](U06){$(\Lb, 6)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U06](U07){$(\Lb, 7)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U07](U08){$(\Lb, 8)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.25) $)
                and ($ (U02.north) + (0,2.25) $) .. (U02.north);
            \draw [thick,>=latex']
            (U03.north) .. controls ($ (U03.north) + (0,2.25) $)
                and ($ (U05.north) + (0,2.25) $) .. (U05.north);
            \draw [thick,>=latex']
            (U04.north) .. controls ($ (U04.north) + (0,4.5) $)
                and ($ (U07.north) + (0,4.5) $) .. (U07.north);
            \draw [thick,>=latex']
            (U05.south) .. controls ($ (U05.south) + (0,-2.5) $)
                and ($ (U08.south) + (0,-2.5) $) .. (U08.south);
            \draw [thick,>=latex']
            (U06.north) .. controls ($ (U06.north) + (0,2.5) $)
                and ($ (U08.north) + (0,2.5) $) .. (U08.north);
        \end{tikzpicture}
        \end{split}
    \end{equation*}

@@

\uncover<3->{%
\begin{block}{Example}
    The word $\La\Lb\Lb\La$ is not a square. Its only perfect matching\vspace{-1.25em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.2,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(\La, 1)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(\Lb, 2)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U02](U03){$(\Lb, 3)$};
            \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(\La, 4)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
                and ($ (U04.north) + (0,2.5) $) .. (U04.north);
            \draw [thick,>=latex']
            (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
                and ($ (U03.north) + (0,1.5) $) .. (U03.north);
        \end{tikzpicture}\vspace{-1.5em}
        \end{split}
    \end{equation*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%o%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Main motivations}
Let $S_n$ be the set of permutations of size $n$.
\bigskip

@@

\uncover<3->{%
To define a product $\cdot : \QQ[C] \otimes \QQ[C] \to \QQ[C]$,
it is some cases more convenient to start by defining a \alert{coproduct}
$\Delta : \QQ[C] \to \QQ[C] \otimes \QQ[C]$.
\medskip}

\uncover<4->{%
Every coproduct fits into the general form
\begin{equation*}
    \Delta(z)
    = \sum_{x, y \in C} \textcolor{RoyalBlue}{\lambda_{x, y}^z} \;
    (x \otimes y)
\end{equation*}
where the $\textcolor{RoyalBlue}{\lambda_{x, y}^z}$ are structure
coefficients of $\QQ$.
\medskip}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Example: unshuffling binary words}
Let $B := \{\La, \Lb\}^*$ be the set of all binary words and consider
the coproduct
\begin{equation*}
    \Delta : \QQ[B] \to \QQ[B] \otimes \QQ[B]
\end{equation*}
on binary words defined by
\begin{equation*}
    \Delta(u) :=
    \sum_{P \sqcup Q = \{1, \dots, |u|\}}
    u_{|P} \otimes u_{|Q}.
\end{equation*}

\uncover<2->{%
\begin{block}{Example}\vspace{-1em}
\begin{equation*}
    \Delta(\Lb\La\La) =
    \epsilon \otimes \Lb\La\La + \Lb \otimes \La\La
    + 2 (\La \otimes \Lb\La) + 2 (\Lb\La \otimes \La) + \La\La \otimes \Lb
    + \Lb\La\La \otimes \epsilon
\end{equation*}
\end{block}}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%o%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Example: shuffling binary words}
\begin{block}{Proposition \BIB{Reutenauer}{1993}}
    The shuffle product of binary words is the dual product of the
    unshuffling coproduct of binary words.
\end{block}
\medskip

\uncover<2->{%
\begin{block}{Example}
Since
\begin{equation*}
    \textcolor{RoyalBlue}{\Delta(\Lb\La\La)} =
    \epsilon \otimes \Lb\La\La + \Lb \otimes \La\La
    + \textcolor{Orange}{2} (\textcolor{BrickRed}{\La \otimes \Lb\La})
    + 2 (\Lb\La \otimes \La) + \La\La \otimes \Lb
    + \Lb\La\La \otimes \epsilon,
\end{equation*}
we observe that $\textcolor{BrickRed}{\La \otimes \Lb\La}$ has
multiplicity $\textcolor{Orange}{2}$ in
$\textcolor{RoyalBlue}{\Delta(\Lb\La\La)}$.
\bigskip

\uncover<3->{%
By duality, this implies that $\textcolor{RoyalBlue}{\Lb\La\La}$ has
multiplicity $\textcolor{Orange}{2}$ in
$\textcolor{BrickRed}{\La \shuffle \Lb\La}$. Indeed,
\begin{equation*}
    \textcolor{BrickRed}{\La \shuffle \Lb\La} =
    \La\Lb\La + \textcolor{Orange}{2}\, \textcolor{RoyalBlue}{\Lb\La\La}.
\end{equation*}}
\vspace{-1.5em}
\end{block}}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Unshuffling permutations}
If $u$ is a word of integers without multiple occurrence of a same letter,
the \alert{standardized} $\STD(u)$ of $u$ is the unique permutation of
$S_{|u|}$ order-isomorphic to~$u$.

\uncover<2->{%
\begin{block}{Example}
\begin{equation*}
    \STD(82194) = 42153
\end{equation*}
\end{block}
\medskip}

\uncover<3->{%
\begin{block}{Definition}
    The \alert{unshuffling coproduct of permutations} is the
    coproduct on $\QQ[S]$ defined by
    \begin{equation*}
        \Delta(\pi) :=
        \sum_{P \sqcup Q = \{1, \dots, |\pi|\}}
        \STD\left(\pi_{|P}\right) \otimes \STD\left(\pi_{|Q}\right).
    \end{equation*}
\end{block}}

>>>>

%\usepackage{euler,palatino}
\usepackage{lmodern}
\usepackage{amsmath,amssymb}
\usepackage[utf8]{inputenc}
\usepackage{xxcolor}
\usepackage{graphicx}
\usepackage[ruled]{algorithm2e}
\usepackage[english]{babel}
\usepackage{rotate}
\usepackage{shuffle}
%\usepackage{minted}
\usepackage{url}
\usepackage{etoolbox}
\usepackage[usestackEOL]{stackengine}
\usepackage{tikz}
\usepackage{rotating}
\usetikzlibrary{snakes}
\usetikzlibrary{trees}
\usetikzlibrary{arrows}
\usetikzlibrary{matrix}
\usetikzlibrary{calc}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[Samuele Giraudo]{
    \textbf{\underline{Samuele~Giraudo}} \qquad
    \textbf{St\'ephane~Vialette}}
\title[Unshuffling Permutations]{
    {\huge \textbf{Unshuffling Permutations}}}
\institute[LIGM]{
    {\footnotesize Laboratoire d'Informatique Gaspard-Monge \\
    Université Paris-Est Marne-la-Vallée \\
    UMR CNRS 8049}}
\date[LATIN 16]{{
    \scriptsize Latin American Theoretical Informatics Symposium \\
    April $11$-$15$ $2016$}}

%\subtitle{
    %\includegraphics[scale=.16]{logos/CNRSfr}
    %\qquad
    %\includegraphics[scale=.10]{logos/ecole_ponts_RVB300.jpg}
    %\qquad
    %\includegraphics[scale=.26]{logos/LIGM}
    %\qquad
    %\raisebox{.25cm}{\includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/LogoEsieeParis}}
    %\qquad
    %\raisebox{.25cm}{\includegraphics[width=1.cm,height=1.25cm,keepaspectratio]{logos/UPEM_LOGO_EDITION}}
%}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Shuffling words}
\begin{block}{{\bf Definition} (shuffle product of words)}
    The \alert{shuffle product} $\shuffle$ on words of $A^*$ is defined
    recursively by \vspace{-.5em}
    \begin{align*}
        u \shuffle \epsilon & := \{u\} =: \epsilon \shuffle u, \\
        (ua \shuffle vb) & := (ua \shuffle v)b \cup (u \shuffle vb)a.
    \end{align*}
\end{block}
\medskip

\uncover<2->{%
To take into account multiplicities, we consider $\shuffle$ as a linear
product
\begin{equation*}
    \shuffle : \QQ[A^*] \otimes \QQ[A^*] \to \QQ[A^*]
\end{equation*}
on $\QQ[A^*]$, the $\QQ$-linear span of words defined by
\begin{align*}
    u \shuffle \epsilon & := u =: \epsilon \shuffle u, \\
    (ua \shuffle vb) & := (ua \shuffle v)b + (u \shuffle vb)a.
\end{align*}}
%
\uncover<3->{%
\begin{block}{\bf Example}
    \vspace{-1em}
    \begin{equation*}\begin{split}
        \textcolor{RoyalBlue}{\La\Lb} \shuffle \textcolor{BrickRed}{\Lb\La}
        & =
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\Lb\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb}%
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb\La}%
        \textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La\Lb}%
        \textcolor{BrickRed}{\La}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La}%
        \textcolor{BrickRed}{\La}\textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb\La}\textcolor{RoyalBlue}{\La\Lb} \\
        &
        \uncover<4->{=
        2 \La\Lb\Lb\La + \La\Lb\La\Lb + \Lb\La\Lb\La + 2 \Lb\La\La\Lb}
    \end{split}\end{equation*}
    \vspace{-1.5em}
\end{block}}
\end{frame}

@@

Given two words $v_1$ and $v_2$, the set $v_1 \shuffle v_2$
can be computed in
\begin{footnotesize}
\begin{equation*}
    O\left((|v_1| + |v_2|) \; \parenVectorstack{|v_1| + |v_2| \\ |v_1|}\right)
\end{equation*}
\end{footnotesize}
time \BIB{Spehner}{1986}.
\bigskip
\bigskip
\bigskip

\uncover<2->{%
Given three words $u$, $v_1$, and $v_2$, deciding if $u$ is in
$v_1 \shuffle v_2$ can be done in
\begin{footnotesize}
\begin{equation*}
    O\left(\frac{|u|^2}{\log(|u|)}\right)
\end{equation*}
\end{footnotesize}
time \BIB{van Leeuwen, Nivat}{1982}.
\bigskip
\bigskip
\bigskip}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Square words}
\begin{block}{{\bf Definition} (square words)}
    A word $u$ is a \alert{square} (w.r.t. the shuffle product
    $\shuffle$) if there is a word $v$ such that $u$ appears
    in~$v \shuffle v$.
\end{block}
\medskip

\uncover<2->{%
\begin{block}{\bf Example}
    The word
    \begin{math}
        u := \textcolor{RoyalBlue}{\Lc}\textcolor{BrickRed}{\Lc}
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\La}
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\Lb\Lb}
    \end{math}
    is a square since $u$ can be obtained by shuffling $\Lc\La\Lb\Lb$
    with itself.
\end{block}
\medskip}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Recognizing square words}
\begin{block}{{\bf Definition} (perfect matchings for words)}
    A \alert{perfect matching} of a word $u$ is a graph $(V, E)$ such
    that
    \begin{itemize}
        \item $V := \{(u_i, i) : i \in \{1, \dots, |u|\}$;
        \item every vertex of $V$ belongs to exactly one edge of $E$;
        \item $(u_i, i) \-- (u_j, j) \in E$ implies $u_i = u_j$.
    \end{itemize}
\end{block}

\uncover<2->{%
\begin{block}{\bf Example}
    The word $\Lb\La\Lb\La\Lb\Lb\Lc\Lc$ admits the perfect matching \vspace{-1em}
    \begin{equation*}
    \begin{split}
    \begin{tikzpicture}[xscale=.5,yscale=.25,inner sep=1pt,node distance=1.5cm]
        \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
        \draw node[Vertex,fill=RoyalBlue!30](U01){$(\Lb, 1)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(\La, 2)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(\Lb, 3)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(\La, 4)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U04](U05) {$(\Lb, 5)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U05](U06){$(\Lb, 6)$};
        \draw node [Vertex,fill=RoyalBlue!30,right of=U06](U07){$(\Lc, 7)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U07](U08){$(\Lc, 8)$};
        \draw [thick,>=latex']
        (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
            and ($ (U06.north) + (0,2.5) $) .. (U06.north);
        \draw [thick,>=latex']
        (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
            and ($ (U04.north) + (0,1.5) $) .. (U04.north);
        \draw [thick,>=latex']
        (U03.south) .. controls ($ (U03.south) + (0,-1.5) $)
            and ($ (U05.south) + (0,-1.5) $) .. (U05.south);
        \draw [thick,>=latex']
        (U07.north) .. controls ($ (U07.north) + (0,1.5) $)
            and ($ (U08.north) + (0,1.5) $) .. (U08.north);
    \end{tikzpicture}
    \end{split}\,.
    \end{equation*}
    \vspace{-1.5em}
\end{block}}

\uncover<3->{%
\begin{block}{{\bf Definition} (inclusion-free perfect matchings)}
    A perfect matching $(V, E)$ of a word $u$ is \alert{inclusion-free}
    if there are no edges $(u_i, i) \-- (u_j, j)$ and
    $(u_k, k) \-- (u_\ell, \ell)$ of $E$ such that $i < k < \ell < j$.
\end{block}}
\end{frame}

@@

\begin{block}{{\bf Proposition} \BIB{Rizzi, Vialette}{2013}}
    A word $u$ is a square iff $u$ admits an inclusion-free perfect
    matching.
\end{block}

\uncover<2->{%
\begin{block}{\bf Example}
    The word
    \begin{math}
        \textcolor{RoyalBlue}{\Lc}\textcolor{BrickRed}{\Lc}
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\La}
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\Lb\Lb}
    \end{math}
    admits the perfect matching \vspace{-1.25em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.16,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(\Lc, 1)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U01](U02) {$(\Lc, 2)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(\La, 3)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U03](U04) {$(\Lb, 4)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U04](U05) {$(\La, 5)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U05](U06){$(\Lb, 6)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U06](U07){$(\Lb, 7)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U07](U08){$(\Lb, 8)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.25) $)
                and ($ (U02.north) + (0,2.25) $) .. (U02.north);
            \draw [thick,>=latex']
            (U03.north) .. controls ($ (U03.north) + (0,2.25) $)
                and ($ (U05.north) + (0,2.25) $) .. (U05.north);
            \draw [thick,>=latex']
            (U04.south) .. controls ($ (U04.south) + (0,-2.25) $)
                and ($ (U07.south) + (0,-2.25) $) .. (U07.south);
            \draw [thick,>=latex']
            (U06.north) .. controls ($ (U06.north) + (0,2.5) $)
                and ($ (U08.north) + (0,2.5) $) .. (U08.north);
        \end{tikzpicture}
        \end{split}
    \end{equation*}

@@

\uncover<3->{%
\begin{block}{\bf Example}
    The word $\La\Lb\Lb\La$ is not a square. Its only perfect matching\vspace{-1em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.2,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(\La, 1)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(\Lb, 2)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U02](U03){$(\Lb, 3)$};
            \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(\La, 4)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2) $)
                and ($ (U04.north) + (0,2) $) .. (U04.north);
            \draw [thick,>=latex']
            (U02.south) .. controls ($ (U02.south) + (0,-1.5) $)
                and ($ (U03.south) + (0,-1.5) $) .. (U03.south);
        \end{tikzpicture}\vspace{-1.5em}
        \end{split}
    \end{equation*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%o%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Main motivations}
Let $S_n$ be the set of \alert{permutations} of size $n$.
\bigskip

@@

\uncover<3->{%
To define a product $\cdot : \QQ[C] \otimes \QQ[C] \to \QQ[C]$,
it is in some cases more convenient to start by defining a \alert{coproduct}
$\Delta : \QQ[C] \to \QQ[C] \otimes \QQ[C]$.
\medskip}

\uncover<4->{%
Every coproduct fits into the general form
\begin{equation*}
    \Delta(z)
    = \sum_{x, y \in C} \textcolor{RoyalBlue}{\lambda_{x, y}^z} \;
    (x \otimes y)
\end{equation*}
where the $\textcolor{RoyalBlue}{\lambda_{x, y}^z} \in \QQ$ are structure
coefficients.
\medskip}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Example: unshuffling words}
Consider the coproduct
\begin{equation*}
    \Delta : \QQ[A^*] \to \QQ[A^*] \otimes \QQ[A^*]
\end{equation*}
on words defined by
\begin{equation*}
    \Delta(u) :=
    \sum_{P \sqcup Q = \{1, \dots, |u|\}}
    u_{|P} \otimes u_{|Q}.
\end{equation*}

\uncover<2->{%
\begin{block}{\bf Example}\vspace{-1.5em}
\begin{equation*}
    \textcolor{RoyalBlue}{\Delta(\Lb\La\La)} =
    \epsilon \otimes \Lb\La\La + \Lb \otimes \La\La
    + \textcolor{Orange}{2} (\textcolor{BrickRed}{\La \otimes \Lb\La})
    + 2 (\Lb\La \otimes \La) + \La\La \otimes \Lb
    + \Lb\La\La \otimes \epsilon
\end{equation*}
\end{block}}

\uncover<3->{%
\begin{block}{{\bf Proposition} \BIB{Reutenauer}{1993}}
    The shuffle product of words is the dual product of the
    unshuffling coproduct of words.
\end{block}}

\uncover<4->{%
\begin{block}{\bf Example}
Since $\textcolor{BrickRed}{\La \otimes \Lb\La}$ has
multiplicity $\textcolor{Orange}{2}$ in
$\textcolor{RoyalBlue}{\Delta(\Lb\La\La)}$, the coeff. of
$\textcolor{RoyalBlue}{\Lb\La\La}$ is
$\textcolor{Orange}{2}$ in
$\textcolor{BrickRed}{\La \shuffle \Lb\La}$:\vspace{-.9em}
\begin{equation*}
    \textcolor{BrickRed}{\La \shuffle \Lb\La} =
    \La\Lb\La + \textcolor{Orange}{2}\, \textcolor{RoyalBlue}{\Lb\La\La}.
\end{equation*}
\vspace{-1.8em}
\end{block}}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Unshuffling permutations}
\begin{block}{{\bf Definition} (standardization of words)}
    If $u$ is a word of integers without multiple occurrence of a same
    letter, the \alert{standardized} $\STD(u)$ of $u$ is the unique
    permutation of $S_{|u|}$ order-isomorphic to~$u$.
\end{block}

\uncover<2->{%
\begin{block}{\bf Example}
\vspace{-.5em}
\begin{equation*}
    \STD(82194) = 42153
\end{equation*}
\end{block}
\medskip}

\uncover<3->{%
\begin{block}{{\bf Definition} (unshuffling coproduct of permutations)}
    The \alert{unshuffling coproduct of permutations} is the
    coproduct on $\QQ[S]$ defined by
    \begin{equation*}
        \Delta(\pi) :=
        \sum_{P \sqcup Q = \{1, \dots, |\pi|\}}
        \STD\left(\pi_{|P}\right) \otimes \STD\left(\pi_{|Q}\right).
    \end{equation*}
\end{block}}

\uncover<4->{%
\begin{block}{\bf Example}
\vspace{-.5em}
\begin{equation*}
    \Delta(213) =
    \epsilon \otimes 213 + 2 (1 \otimes 12) +
    1 \otimes 21 + 2 (12 \otimes 1) + 21 \otimes 1
    + 213 \otimes \epsilon
\end{equation*}
\end{block}}
\end{frame}
